---
title: "GEOG 595 rayshader Activity by Jasper Zhou"
output: html_notebook
---

**Let's check if the package, rayshader, is installed.**
```{r}
# Function to check and install rayshader if not installed
check_install_rayshader <- function() {
  if (!requireNamespace("rayshader", quietly = TRUE)) {
    install.packages("rayshader", dependencies = TRUE, quiet = TRUE)
  }
}

check_install_rayshader()
```

**Import the required libraries.**
```{r}
library(rayshader)
library(terra)
library(rgl)
```

**Where can you obtain DEM data?**
1. USGS Earth Explorer
2. Open Topography
3. National Map Viewer
...

Note: Highest resolution DEM may be available from state agency. For instance, Washington State Department of Natural Resources (DNR) provides 1m resolution DEMs for free.

```{r}
# Download & Load the DEM from GitHub
dem_url <- "https://github.com/Fangsheng-Zhou/rayshader_exploration/raw/refs/heads/main/Example_DEM_Data/USGS_13_n49w122_20230307_SCG.tif"
dest_file <- "Data/scg_dem.tif"

# Download and save the DEM
download.file(dem_url, destfile = dest_file, mode = "wb")

# Read the DEM as a raster
scg_dem <- rast(dest_file)

# Plot the DEM as a grayscale image to verify
plot(scg_dem, main = "South Cascade Glacier DEM")
```

**Create a 2D plot of the DEM using rayshader.**
```{r}
# Convert to rayshader-compatible matrix
elmat <- raster_to_matrix(scg_dem)

# Apply shading and plot the 2D rayshader map
elmat %>%
  sphere_shade(texture = "desert") %>%
  plot_map()
```

**Adding some shadow**
```{r}
# Apply texture and simple ray-traced shadows
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>% # Adds realistic lighting effects
  plot_map()
```

**Plot it in 3D**
```{r}
# Render in 3D 
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>% # Adds realistic lighting effects
  plot_3d(elmat, zscale = 10, fov = 0, theta = 135, zoom = 0.75, phi = 45, windowsize = c(1000, 800))

# Pause briefly before taking a snapshot
Sys.sleep(3)

# Save a high-quality snapshot
render_snapshot("Data/SCG_3D.png")
```

**Create a 3D animation**
Note: Don't close the rgl window until it finished rendering.
```{r}
# Render in 3D 
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>% # Adds realistic lighting effects
  plot_3d(elmat, zscale = 10, fov = 0, theta = 135, zoom = 0.75, phi = 45, windowsize = c(1000, 800))

# Pause briefly before taking a snapshot
Sys.sleep(3)

render_movie(
  filename = "Data/SCG_3D_animation.mp4",  # Save as .mp4
  frames = 180,   # Number of frames (adjust for smoother animation)
  phi = 45,       # Camera angle
  theta = seq(0, 360, length.out = 180),  # Rotate 360 degrees
  zoom = 0.75
)
```

**Add in a scale bar**
```{r}
# Render in 3D 
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>% # Adds realistic lighting effects
  plot_3d(elmat, zscale = 10, fov = 0, theta = 135, zoom = 0.75, phi = 45, windowsize = c(1000, 800))

render_camera(fov = 0, theta = 200, zoom = 0.85, phi = 45)

# Add a scale bar
render_scalebar(limits=c(0, 500, 1000),label_unit = "m",position = "W", y=50,
                scale_length = c(0.33,1))
```

**Add in a compass after scale bar**
```{r}
# Render in 3D 
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>% # Adds realistic lighting effects
  plot_3d(elmat, zscale = 10, fov = 0, theta = 135, zoom = 0.75, phi = 45, windowsize = c(1000, 800))

render_camera(fov = 0, theta = 200, zoom = 0.85, phi = 45)

# Add a scale bar
render_scalebar(limits=c(0, 500, 1000),label_unit = "m",position = "W", y=50,
                scale_length = c(0.33,1))

# Add a compass
render_compass(position = "E")

Sys.sleep(1)

# Save a high-quality snapshot
render_snapshot("Data/SCG_3D_scale_compass.png")
```

**Getting little artsy - depth of field**
Note: This will take a long time to render. 
```{r}
# Render in 3D 
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>% # Adds realistic lighting effects
  plot_3d(elmat, zscale = 10, fov = 0, theta = 135, zoom = 0.75, phi = 45, windowsize = c(1000, 800))

render_camera(fov = 0, theta = 200, zoom = 0.45, phi = 45)

render_depth(focallength = 600, clear = TRUE)

Sys.sleep(1)

# Save a high-quality snapshot
render_snapshot("Data/SCG_3D_scale_DOF.png")
```

**Let's do a 3D plot.**
```{r}
library(ggplot2)
library(rayshader)
```

**2D vs. 3D plots, side by side comparison**
```{r}
# Exmaple 1
# Use the built-in volcano dataset
data(volcano)

# Convert the volcano matrix to a data frame for ggplot2
volcano_df <- reshape2::melt(volcano)

# Create a ggplot2 heatmap
gg <- ggplot(volcano_df, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradientn(colors = terrain.colors(10)) +
  theme_minimal() +
  labs(title = "Volcano Topography", x = "X", y = "Y", fill = "Height")

# Display the ggplot2 plot
plot(gg)

# Convert the ggplot2 plot to a 3D plot
plot_gg(gg, width = 5, height = 5, scale = 300, multicore = TRUE, windowsize = c(1000, 800))
```

```{r}
# Example 2
# Use the built-in airquality dataset
data(airquality)

# Create a ggplot2 scatter plot
gg <- ggplot(airquality, aes(x = Temp, y = Wind, color = Ozone)) +
  geom_point(size = 3) +  # Scatter plot points
  scale_color_gradientn(colors = heat.colors(10)) +  # Color gradient
  theme_minimal() +
  labs(title = "Airquality Dataset: Temperature vs Wind", 
       x = "Temperature", y = "Wind", color = "Ozone")

# Display the 2D ggplot
print(gg)

# Convert the ggplot2 plot to a 3D plot
plot_gg(gg, width = 5, height = 5, scale = 300, multicore = TRUE, windowsize = c(1000, 800))
```

**Some cool things you can do with rayshader**
1. https://opentopography.org/blog/creating-2d-and-3d-visualizations-rayshader
2. https://rallydatajunkie.com/visualising-rally-stages/index.html

**References**
Morgan-Wall T (2025). rayshader: Create Maps and Visualize Data in 2D and 3D. R package version 0.38.11, https://github.com/tylermorganwall/rayshader, https://www.rayshader.com.